---
title: "Numerical EDA"
output: github_document
---

```{r setup, include = FALSE}
library(tidyverse)
library(p8105.datasets)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

Load the weather data as well.

```{r}
data("weather_df")

weather_df = 
  weather_df |>
  mutate(
    month = lubridate::floor_date(date, unit = "month"))
```

Now we have everything we need!

```{r}
weather_df |> 
  filter(prcp>0) |> 
  ggplot(aes(x=prcp)) +
  geom_histogram()
```

```{r}
weather_df |> 
  filter(prcp>1000)
```

```{r}
weather_df |> 
  filter(tmax >= 20, tmax <= 30) |> 
  ggplot(aes(x = tmin, y = tmax, color = name, shape = name)) + 
  geom_point(alpha = .75)
```
```{r}
weather_df |>
  group_by(name, month)
```


```{r}
weather_df |> 
  group_by(month) |> 
  summarize(
    count=n(),
    n_days = n_distinct(date))
```

DONT DO THIS
```{r, eval = FALSE}
weather_df |> 
  pull(tmax) |> 
  summary()
```

do this instead

```{r}
weather_df |> 
  group_by(name) |> 
  summarize(
    n = n(),
    mean_tmax = mean(tmax, na.rm = TRUE),
    median_tmax = median(tmax, na.rm= TRUE),
    sd_prcp = sd(prcp, na.rm = TRUE),
    q25_prcp = quantile(prcp, 0.25, na.rm = TRUE),
    q95_prcp = quantile(prcp, 0.95, na.rm = TRUE)
  ) |> 
  knitr::kable(digits = 2)
```

```{r}
weather_df |> 
  group_by(name, month) |> 
  summarise(
    mean_tmax = mean(tmax, na.rm = TRUE) #removes NAs from mean calculation
  ) |> 
  pivot_wider(
    names_from = name,
    values_from = mean_tmax
  ) |> 
  knitr::kable(digits = 2)
```

what abut looking at this
```{r}
weather_df |> 
  group_by(name, month) |> 
  summarise(
    mean_tmax = mean(tmax, na.rm = TRUE)
  ) |> 
  ggplot(aes(x = month, y = mean_tmax, color = name)) +
  geom_point() +
  geom_line()
```
```{r}
weather_df |> 
  mutate(center_tmax = tmax - mean(tmax, na.rm = TRUE)) |> 
  ggplot(aes(x = date, y = center_tmax, color = name))+
  geom_point()
```

What about "window" functions?

Try to rank things

```{r}
weather_df |> 
  group_by(name, month) |> 
  mutate(temp_rank = min_rank(desc(tmax))) |> # desc = descending order; min_rank handles ties differently (gives the min)
  filter(temp_rank < 2)
```

lead and lag

```{r}
weather_df |> 
  group_by(name) |> 
  mutate(
    lagged_tmax = lag(tmax, 3), #need the group_by for this to make sense; lag by 3 days
    lead_tmax = lead(tmax)
  )
```

```{r}
weather_df |> 
  group_by(name) |> 
  mutate(
    temp_change = tmax - lag(tmax)
  ) |> 
  summarize(
    mean_temp_change = mean(temp_change, na.rm = TRUE),
    sd_temp_change = sd(temp_change, na.rm = TRUE)
  )
```

## Revisit some examples

import, clean, tidy, etc pulse data and compute the mean and median BDI score at each visit
```{r}
pulse_df = haven::read_sas("data/public_pulse_data.sas7bdat") |> 
  janitor::clean_names() |> 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    names_prefix = "bdi_score_",
    values_to = "bdi_score"
  ) |> 
  select(id, visit, everything()) |> 
  mutate(
    visit = replace(visit, visit == "bl", "00m")
  )

pulse_df |> 
  group_by(visit) |> 
  summarize(
    mean_bdi_score = mean(bdi_score, na.rm = TRUE),
    median_bdi_score = median(bdi_score, na.rm = TRUE)
  ) |> 
  knitr::kable(digits=2)
```

in the FAS data, compute mean outcome (ears only) across dose and day of treatment; show in a reader-friendly table
```{r}
pups_df = read_csv("data/FAS_pups.csv", skip = 3, na = c("", ".", "NA")) |> 
  janitor::clean_names()

litters_df = read_csv("data/FAS_litters.csv", na = c("", ".", "NA")) |> 
  janitor::clean_names() |> 
  separate(group, into = c("dose", "day_of_tx"), 3)

fas_df = 
  left_join(pups_df, litters_df, by = "litter_number") |> 
  select(litter_number, dose, day_of_tx, everything()) |> 
  drop_na(dose, day_of_tx)

fas_df |> 
  group_by(dose, day_of_tx) |> 
  summarize(
    mean_ears = mean(pd_ears, na.rm = TRUE)
  ) |> 
  pivot_wider(
    names_from = day_of_tx, 
    values_from = mean_ears
  )
```

